Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

### 2024-06-14 EBDVF-2024/paper.qmd
Update paper.qmd

abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated data.
  The Vocabulary Hub service (part of the IDSA RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.

@@

# Projects and Use Cases
In the expanding data economy landscape supported by dataspaces, 
we are working on several projects which address practical data sharing needs and were we experience the importance of semantic interoperability to fulfill various use cases. 
our main focus to develop our solution is the [UNDERPIN](https://underpinproject.eu/) project with use cases of refineries and wind farms,
but also other project, like [DataBri-X](https://databri-x.eu/) provide use cases which were considered.

## Manufacturing/Maintenance Dataspace
Operating refineries and wind farms involves high maintenance costs.
Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business-critical.
For establishing predictive maintenance, we need high amounts of high quality training data
that can be obtained by data consolidation across different refinery and wind farm measurement sources.

The UNDERPIN project addresses this problem by creating a dataspace for refineries, wind farms, and their equipment suppliers.
Extensive time series data is exchanged, but their format and meaning is not standardized.
Describing the meaning of these time series facilitates semantic interoperability between datasets from different providers.
This description includes measured variable, quantity kind and unit of measure,
which sensor took the measurements, what it is attached to, the detailed model and connectivity of that industrial equipment.

@@

### The Wind Farm Dataspace
Operating wind farms involves high costs for maintaining the wind turbines.
Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business critical.
For establishing predictive maintenance, we need high amounts of high qualitative training data,
which can be achieved by consolidation different wind farm measurement sources.
This use case provides different kinds of tabular data, where consolidation needs semantic harmonization of data.

@@

## IDS Vocabulary Hub
In this paper, we focus on the IDSA approach to dataspaces.
IDSA defines the architecture of dataspaces in the reference architecture model IDS-RAM (current version 4).
One of the defined architectural components, the Vocabulary Hub, has the role to provide vocabularies to annotate and describe data assets and services.
These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions. 
It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies 
and thereby provides semantic interoperability in a standardized and machine-readable way.

@@

Both GraphDb and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL. They therefore seamlessly integrate with each other, but also with other services and components in dataspaces which support these machine-readable standards for semantic interoperability.

# A Semantic Solution and its Benefits
When harmonizing data, we need to provide a method to automatically map and process it,
so this can be done efficiently on large datasets and/or regular updates in dataspaces. 
The benefit provided should support the ML training in such a way that the consolidated dataset is exposed as uniform training dataset.

@@

- A Data Provider sends data through the Connector, 
  and at the same time provides basic metadata (DCAT)
- Selected data can be stored in Database(s) managed by the dataspace,
  for easier access and querying.
  - For large-volume time series data, we recommend to use InfluxDB, PostgreSQL, Google BigQuery, etc.
  - Of course, access control is very important here.
    We'll leverage GraphDB's [Fine-grained Access Control](https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html), 
    but access control in the time series database is also important
  - Depending on Provider preferences and agreements, 
    other data may stay at the source, 
    and be delivered to the Consumer only dynamically through the Connector
- The Vocabulary Hub stores relevant semantic assets: ontologies and thesauri.
  These are added based on relevant use cases and datasets, following a dataspace governance process.
- The Vocabulary Hub also stores mapping assets, such as OBDA or R2RML mappings and CSVW manifests.
- PoolParty Semantic Integrator is used to edit, view and manage the semantic assets,
  and to bind mapping assets to incoming data.
- For virtual access to relational data, ONTOP is used to translate SPARQL to SQL dynamically
  and to convert returned data to SPARQL result format.
- The Data Consumer benefits from:
  - More powerful dataset discovery by using richer semantic queries, eg
    - "Give me all time series related to compressor C123" (equipment instance)
    - "Give me all time series related to temperature" (measured quantity)
    - "Give me all time series of bearings" (equipment kind or machine part)
  - Harmonized querying of datasets from multiple providers.

>>>>

abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
  The Vocabulary Hub service (part of the IDSA RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.

@@

# Projects and Use Cases
In the expanding data economy landscape supported by dataspaces, 
we are working on several projects which address practical data sharing needs and where we experience the importance of semantic interoperability to fulfill various use cases. 
Our solution is developed with main focus on the [UNDERPIN](https://underpinproject.eu/) project with use cases of refineries and wind farms, 
but other projects, such as [DataBri-X](https://databri-x.eu/) provide use cases which were considered.

## Manufacturing/Maintenance Dataspace
Operating refineries and wind farms involves high maintenance costs. 
Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business-critical. 
For establishing predictive maintenance, we need large amounts of high quality training data 
that can be obtained by data consolidation across different refinery and wind farm measurement sources.

The UNDERPIN project addresses this problem by creating a dataspace for European manufacturers in the refinery and renewable energy domains and various SMEs across their value-chain ecosystem, such as equipment suppliers. 
Extensive time series data is exchanged, but their format and meaning is not standardized. 
Describing the meaning of these time series facilitates semantic interoperability between datasets from different providers. 
This description includes measured variable, quantity kind and unit of measure, 
which sensor took the measurements, what it is attached to, the detailed model and connectivity of that industrial equipment.

@@

### The Wind Farm Dataspace
Operating wind farms involves high costs for maintaining the wind turbines. 
Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business critical. 
For establishing predictive maintenance, we need large amounts of high qualitative training data, 
which can be achieved by consolidating different wind farm measurement sources. 
This use case provides different kinds of tabular data, where consolidation needs semantic harmonization of data.

@@

## IDS Vocabulary Hub
In this paper, we focus on the IDSA approach to dataspaces. 
IDSA defines the architecture of dataspaces in the reference architecture model IDS-RAM (current version 4). 
One of the defined architectural components, the Vocabulary Hub, has the role of providing vocabularies to annotate and describe data assets and services.
These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions. 
It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies 
and thereby provides semantic interoperability in a standardized and machine-readable way.

@@

Both GraphDB and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL. They therefore seamlessly integrate with each other, but also with other services and components in dataspaces which support these machine-readable standards for semantic interoperability.

# A Semantic Solution and its Benefits
When harmonizing data, we need to provide a method to automatically map and process it, 
so this can be done efficiently on large datasets and/or regular updates in dataspaces. 
The benefit provided should support the ML training in such a way that the consolidated dataset is exposed as a uniform training dataset.

@@

- A Data Provider sends data through the Connector, 
  and at the same time provides basic metadata (DCAT)
- Selected data can be stored in Database(s) managed by the dataspace, 
  for easier access and querying.
  - For large-volume time series data, we recommend to use InfluxDB, PostgreSQL, Google BigQuery, etc.
  - Of course, access control is very important here. 
    We'll leverage GraphDB's [Fine-grained Access Control](https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html), 
    but access control in the time series database is also important.
  - Depending on Provider preferences and agreements, 
    other data may stay at the source, 
    and be delivered to the Consumer only dynamically through the Connector.
- The Vocabulary Hub stores relevant semantic assets: ontologies and thesauri. 
  These are added based on relevant use cases and datasets, following a dataspace governance process.
- The Vocabulary Hub also stores mapping assets, such as OBDA or R2RML mappings and CSVW manifests.
- PoolParty Semantic Integrator is used to edit, view and manage the semantic assets, 
  and to bind mapping assets to incoming data.
- For virtual access to relational data, ONTOP is used to translate SPARQL to SQL dynamically 
  and to convert returned data to SPARQL result format.
- The Data Consumer benefits from:
  - More powerful dataset discovery by using richer semantic queries, e.g.
    - "Give me all time series related to compressor C123" (equipment instance)
    - "Give me all time series related to temperature" (measured quantity)
    - "Give me all time series of bearings" (equipment kind or machine part)
  - Harmonized querying of datasets from multiple providers.
